### 2022-09-20 AISTATS2023/paper.tex
added bin spliting script

The experimental section is divided in simulation and real-world
examples. The simulation examples, where the data-generating
distribution \(p(\mathbf{X})\) and the predictive function
\(f(\cdot)\) are defined by us, are used for comparing all methods
against a solid ground-truth. Having the freedom to experiment with
\(p(\mathbf{X}), f(\cdot)\), we explore the behavior of each method in
different settings for highlighting important aspects
(advantages/disadvantages) of each one. In particular, we design two
end-to-end simulation examples; At the first one, we focus on
comparing NAME method versus PDP-ICE in terms of quantifying the
heterogeneous effects for a range of different black-box functions. At
the second one, we provide empirical clues that NAME method achieves a
better approximation compared to traditional ALE, due to appropriate
bin-splitting. Afterwards, we also apply the methods on some
real-world datasets. In these cases, it is infeasible to obtain ground
truth information, therefore, we perform evaluation at two
levels. First, we compare visually NAME versus PDP with ICE
plots. Second, we hold as ground-truth the effects computed on the
full training-set and we, then, perform subsampling to evaluate the
robustness of the bin-splitting method.

@@

In general, there is no a single unchallenged aggreement about the
correct feature effect; each method models it in a slightly different
way. Therefore, for creating a universal evaluation framework we
reside on variations of the simple piecewise-linear function, where in
most cases, we can have an homophony about the correct effect.

In all the experiments of the example, we will use the following
generative distribution: \(p(\mathbf{x}) = p(x_3)p(x_2|x_1)p(x_1)\)
where \(x_1 \sim \mathcal{U}(0,1)\),
\(x_2 \sim \mathcal{N}(x_1, \sigma_2^2)\),
\(x_3 \sim \mathcal{N}(0, \sigma_3^2)\). 
%
The predictive function is:

\begin{equation}
  \label{eq:2}
  f(\mathbf{x}) = \begin{cases}
                    f_{\mathtt{lin}} + \alpha f_{\mathtt{int}} & \text{if $f_{\mathtt{lin}} < 0.5$ }\\
                    0.5 - f_{\mathtt{lin}} + \alpha f_{\mathtt{int}} & \text{if $0.5 \leq f_{\mathtt{lin}} < 1$}\\
                    \alpha f_{\mathtt{int}} &\text{otherwise}
                  \end{cases}
\end{equation}
%
where \(f_{\mathtt{lin}}(\mathbf{x}) = a_1 x_1 + a_2 x_2\) and
\(f_{inter}(\mathbf{x}) = x_1x_3\). The term \(f_{\mathtt{lin}}\) is a
linear function with \(a_1, a_2\) the weights of the corresponding
variables. The term \(f_{\mathtt{lin}}\) is the interaction between
the independent variables \(x_1\) and \(x_3\), with \(\alpha\) setting
the weight of the interaction term.

\subsubsection{Effect Without Uncertainty}

We test the feature effect when no interaction term is apparent,
i.e. \(\alpha=0\), there is no uncertainty about the value of \(x_2\),
i.e. \(x_2 = x_1\) and the weights are equal weights \(a_1=a_2=1\).
In this case, ground truth effect for \(f^{\mathtt{GT}}(x_1)\) is
\(x_1\) when \(0 \leq x_1 < 0.25\), \(-x_1\) when
\(0.25 \leq x_1 < 0.5\) and zero otherwise. In each position \(x_1\),
there is complete absence of heterogeneous effects, and therefore, the
uncertainty of the global explanation is zero,
i.e. \(f^{\mathcal{GT}}_{\sigma^2}(x_1) = 0\). In figure we present,
the effect of each method;

\begin{figure}[h]
  \centering
  \includegraphics[width=.23\textwidth]{example_3/dale_auto_bins.png}
  \includegraphics[width=.23\textwidth]{example_3/pdp_ice.png}
  \caption{Figure 1}
  \label{fig:ex-synth-1-1}
\end{figure}

\subsubsection{Effect With Uncertainty}

@@

The perfect bin-splitting according to the definition of the
minimization problem, is the one that creates:

>>>>

The experimental section is divided in simulation and real-world
examples. The simulation examples, where the data-generating
distribution \(p(\mathbf{X})\) and the predictive function
\(f(\cdot)\) are defined by us, are used for comparing all methods
against a solid ground-truth. In particular, we design two end-to-end
simulation examples; At the first one, we focus on comparing NAME
method versus PDP-ICE in terms of quantifying the heterogeneous
effects for a range of different black-box functions. At the second
one, we provide empirical clues that NAME method achieves a better
approximation compared to traditional ALE, due to appropriate
bin-splitting. Afterwards, we also apply the methods on some
real-world datasets. In these cases, it is infeasible to obtain ground
truth information, therefore, we perform evaluation at two
levels. First, we compare visually NAME versus PDP with ICE
plots. Second, we hold as ground-truth the effects computed on the
full training-set and we, then, perform subsampling to evaluate the
robustness of the bin-splitting method.

@@

This synthetic example aims to showcase the superiority of NAME method
in quantifying the main effect and the uncertainty compared to PDP
with ICE plots. For being able to compare the two methods against a
solid ground-truth, we limit ourserlves a simple evaluation framework based
on following piecewise linear function:

\begin{equation}
  \label{eq:synth-ex-1-function}
  f(\mathbf{x}) = \begin{cases}
                    f_{\mathtt{lin}} + \alpha f_{\mathtt{int}} & \text{if $f_{\mathtt{lin}} < 0.5$ }\\
                    0.5 - f_{\mathtt{lin}} + \alpha f_{\mathtt{int}} & \text{if $0.5 \leq f_{\mathtt{lin}} < 1$}\\
                    \alpha f_{\mathtt{int}} &\text{otherwise}
                  \end{cases}
\end{equation}
%
where \(f_{\mathtt{lin}}(\mathbf{x}) = a_1 x_1 + a_2 x_2\) and
\(f_{int}(\mathbf{x}) = x_1x_3\). As we observe, the linear term
\(f_{\mathtt{lin}}\) includes the two correlated features and the term
\(f_{\mathtt{int}}\) interacts the two non-correlated variables. The
samples that we use in all examples are coming from the following
distribution: \(p(\mathbf{x}) = p(x_3)p(x_2|x_1)p(x_1)\) where
\(x_1 \sim \mathcal{U}(0,1)\),
\(x_2 \sim \mathcal{N}(x_1, \sigma_2^2=0)\) and
\(x_3 \sim \mathcal{N}(0, \sigma_3^2)\). We will test three variations
by changing the parameters of the black-box function and of the
generative distribution. Despite the model's simplicity, as we will
see below, PDP with ICE plot fail completely to capture the
ground-truth effect, due to the correlation between features \(X_1\)
and \(X_2\). Furthermore, besides not the focus of the current
example, NAME method manages to automatically extract the regions with
constant effect. Details for obtaining the ground truth effects and
the experimental set-up are in the Appendix.\footnote{The Appendix
  contains the effects for all the features, some of them are ommited
  here for compact presentation.}

\paragraph{No Interaction, Equal weights.}

We test the feature effect when no interaction term is apparent,
i.e. \(\alpha=0\), and the weights are equal \(a_1=a_2=1\).  In this
case, ground truth effect for \(f^{\mathtt{GT}}(x_1)\) is \(x_1\) when
\(0 \leq x_1 < 0.25\), \(-x_1\) when \(0.25 \leq x_1 < 0.5\) and zero
otherwise. In each position \(x_1\), there is complete absence of
heterogeneous effects, and therefore, the uncertainty of the global
explanation is zero, i.e. \(f^{\mathcal{GT}}_{\sigma^2}(x_1) = 0\). In
Figure~\ref{fig:ex-synth-1-1} we see that PDP effect is completely
wrong and ICE plots imply heterogeneous effects. In contrast, ALE
captures correctly the average effect and the zero uncertainty.

\begin{figure}[h]
  \centering
  \includegraphics[width=.23\textwidth]{example_1/dale_feat_0.png}
  \includegraphics[width=.23\textwidth]{example_1/pdp_ice_feat_0.png}
  \caption{Feature effect for \(x_1\); NAME method at the left and PDP
    with ICE at the right.}
  \label{fig:synth-ex-1-case-1}
\end{figure}

\paragraph{No Uncertainty, Different weights.}

As before, there is no interaction term, i.e. \(\alpha=0\), and
uncertainty about the value of \(x_2\), i.e.  \(x_2 = x_1\), but we
give different weight to each feture \(a_1=2, a_2=0.5\). Again, we
observe that ALE quantifies perfectly the ground truth effect and the
zero-uncertainty in contrast to PDP with ICE.

\begin{figure}[h]
  \centering
  \includegraphics[width=.23\textwidth]{example_2/dale_feat_0.png}
  \includegraphics[width=.23\textwidth]{example_2/pdp_ice_feat_0.png}
  \caption{Figure 1}
  \label{fig:ex-synth-1-1}
\end{figure}

\paragraph{Uncertainty, Equal weights.}

In this case we activate the interaction term, i.e. \(a=1\), and we
set equal weights \(a_1=a_2=1\). In contrast with above, the
interaction term provokes heterogeneous effects for features
\(x_1, x_3\); at any position \(x_1\), the local effects dependend on
the unknown value of \(X_3\) and vice-versa. The effect of \(x_2\)
continues to have zero-uncertainty, since it does not appear in any
interaction term. NAME captures perfectly the effect of all features,
and the uncertainty in the case of \(x_1, x_2\). On the contrary,
PDP-ICE quantify the effect and the uncertainty correctly only in the
case of \(x_3\). For all other features, both the average effect and
the uncertainty implied by ICE plots are wrong.

\begin{figure}[h]
  \centering
  \includegraphics[width=.23\textwidth]{example_3/dale_feat_0.png}
  \includegraphics[width=.23\textwidth]{example_3/pdp_ice_feat_0.png}\\
  \includegraphics[width=.23\textwidth]{example_3/dale_feat_1.png}
  \includegraphics[width=.23\textwidth]{example_3/pdp_ice_feat_1.png}\\
  \includegraphics[width=.23\textwidth]{example_3/dale_feat_2.png}
  \includegraphics[width=.23\textwidth]{example_3/pdp_ice_feat_2.png}\\
  \caption{Figure 1}
  \label{fig:ex-synth-1-1}
\end{figure}

\paragraph{Discussion.}

The examples presented above, do not cover all cases. Therefore, they
are not sufficiect to prove that NAME method quantifies the effect
correctly in all cases. Their main objective is to highlight that PDP
indifference for correlated features provoques erroneous effects, even
in very simple cases. Unfortunately, the methods that are based on
PDP and ICE to quantify heterogeneity are also vulnerable to the same
errors.

The examples above do not cover the case where the interaction term
include correlated features, \(x_1, x_2\) for our example. In this
case, there is controversy about the ground-truth
effect~\citep{Gromping2020MAEP}.

@@

The aim of this example is to illustrate the advantages of automatic
bin-splitting based on the objective of Eq.(\ref{eq:opt}). We explore
different setups (black-box functios and dististributions) where, in
each one, we treat as ground-truth the ALE obtained with very dense
sampling. Afterwards, we measure the error of the ALE approximation
with \(K\) fixed-size bins versus the approximation of NAME with
automatic bin-spltting.
